---
title: "Start-to-end analysis with aRchiteutis"
author: "aRchiteutis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Start-to-end analysis with aRchiteutis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  dev = "png"
)

# adiv pulls in rgl, which must run headless in a vignette build.
Sys.setenv(RGL_USE_NULL = "TRUE")
options(rgl.useNULL = TRUE)
set.seed(1)

library(aRchiteutis)
library(ggplot2)

# LIGHT theme + viridis everywhere. ggpubr is a Suggests dependency; when it is
# available we adopt its publication-ready light theme as the global default,
# otherwise we fall back to theme_light(). Every custom chunk below layers a
# viridis palette on top so we never fall back to ggplot2's default hues.
have_ggpubr <- requireNamespace("ggpubr", quietly = TRUE)
if (have_ggpubr) {
  ggplot2::theme_set(ggpubr::theme_pubr())
} else {
  ggplot2::theme_set(ggplot2::theme_light())
}
```

`aRchiteutis` (formerly the *samovar* collection of scripts) is a toolkit for
parsing, transforming and visualising the taxonomic-composition tables produced
by [Kraken2](https://ccb.jhu.edu/software/kraken2/) (and Kaiju / Bracken)
metagenomic classifiers. This vignette walks through a complete analysis of the
example honey-bee gut dataset shipped with the package: from reading the raw
classifier reports, through composition, alpha- and beta-diversity, ordination
and differential abundance, all the way to a co-occurrence network — with a
statistical comparison of the two developmental stages (larvae vs pupa) at each
step.

Throughout we keep the styling consistent: a light theme
(`ggpubr::theme_pubr()`), the colour-blind-friendly **viridis** palette for all
colours and fills, and [`ggpubr`](https://rpkgs.datanovia.com/ggpubr/) for the
statistical layers (`stat_compare_means()` significance annotations).

# 1. Load and inspect the data

The package ships 17 Kraken2 bacterial reports in `inst/extdata/`, one per
sample, plus a `legend.csv` describing each sample's developmental `stage`
(larvae / pupa) and `hive` (1 / 2). We read them all with `get_counts()`,
splitting each file name on `"_"` so the sample id becomes e.g. `m11`, and
joining the legend on.

```{r load}
path <- system.file("extdata", package = "aRchiteutis")
legend <- system.file("extdata", "legend.csv", package = "aRchiteutis")

df <- get_counts(
  path = path,
  pattern = "_decont_bacteria_k2",
  legend = legend,
  trim_char = "_"
)

df
```

The result is one long (tidy) row per taxon per sample. The first six columns
have fixed positions — `taxa`, `clade`, `sample`, `N` (reads), `amount`
(fraction of all reads) and `amount_cl` (fraction of classified reads) —
followed by the legend columns (`stage` in column 7, `hive` in column 8), which
downstream functions reference by index.

```{r inspect}
# Sample metadata (one row per sample).
meta <- unique(df[, c("sample", "stage", "hive")])
meta

# How many samples per stage?
table(meta$stage)

# Number of distinct taxa recorded at each Kraken clade level.
tapply(df$taxa, df$clade, function(x) length(unique(x)))

# Total classified reads per stage (root rows, clade "R").
aggregate(N ~ stage, data = df[df$clade == "R", ], FUN = sum)
```

# 2. Transform the table

The `df_*` helpers reshape the tidy table. The most common operations are:
keeping the most abundant taxa, dropping unclassified rows, and rescaling
per-sample amounts back to one.

```{r transforms}
# Genus-level view, top 8 genera by mean abundance (rolls the rest into
# "other <domain>"), unclassified rows dropped.
df_top <- df_taxa_trim(df[df$clade != "S", ], top_taxa = 8)

# Drop unclassified taxa and rescale amounts to sum to one per sample.
df_clean <- df_rescale(df_tidy_drop_unclassified(df))

# A taxa-by-sample matrix of the top 12 genera, for the matrix-oriented plots.
mat_g <- df_untidy(df, clade = "G", top = 12)
dim(mat_g)
```

# 3. Composition

Three complementary views of *what is there*. `df2donut()` shows the mean
composition across all samples, `df2composition()` stacks each sample, and
`df2barplot()` shows the per-taxon distribution. All three already use viridis
internally.

```{r composition-donut, fig.height = 5}
df2donut(df_top)
```

```{r composition-stack, fig.height = 5}
df2composition(df_top)
```

```{r composition-box, fig.height = 5}
df2barplot(df_top)
```

`df2barplot()` (and the alpha-diversity plots) can also draw a combined
half-violin / half-boxplot via the optional
[`ggviolinbox`](https://github.com/dsmutin/ggviolinbox) package. The chunk below
only runs when that package is installed.

```{r composition-violinbox, eval = requireNamespace("ggviolinbox", quietly = TRUE), fig.height = 5}
df2barplot(df_top, violinbox = TRUE)
```

# 4. Alpha diversity

`df2alpha_summary()` computes a broad panel of alpha-diversity indices per
sample and shows them split by a legend variable; `df2alpha()` does the same for
a chosen shortlist. Here we split by `stage` (column 7) and colour by `hive`
(column 8).

```{r alpha-panel, fig.height = 6}
df_sp <- df[df$clade == "S", ]
df2alpha(df_sp, split_by = 7, add_legend = 8)
```

## 4.1 Statistics: larvae vs pupa

To formally compare the two stages we compute the Shannon and Simpson indices
per sample (with `abdiv`) and test the group difference with a Wilcoxon
rank-sum test, annotated directly on the plot with
`ggpubr::stat_compare_means()`. Fills use viridis and the theme is light.

```{r alpha-table}
alpha_tab <- df_sp %>%
  dplyr::group_by(sample) %>%
  dplyr::summarise(
    Shannon = abdiv::shannon(N),
    Simpson = abdiv::simpson(N),
    .groups = "drop"
  )
alpha_tab <- dplyr::left_join(alpha_tab, meta, by = "sample")
alpha_long <- tidyr::pivot_longer(alpha_tab, c("Shannon", "Simpson"),
                                  names_to = "index", values_to = "value")
head(alpha_tab)
```

```{r alpha-stats, eval = requireNamespace("ggpubr", quietly = TRUE), fig.height = 4.5}
ggpubr::ggviolin(
  alpha_long, x = "stage", y = "value", fill = "stage",
  palette = viridis::viridis(2), alpha = 0.7,
  add = "jitter", add.params = list(color = "grey30", size = 1.5),
  facet.by = "index", scales = "free_y"
) +
  ggpubr::stat_compare_means(
    ggplot2::aes(label = ggplot2::after_stat(p.format)),
    method = "wilcox.test", label.x.npc = "center"
  ) +
  labs(x = NULL, y = "Diversity index")
```

The p-value printed above each panel is the Wilcoxon test comparing larvae with
pupa for that index.

# 5. Beta diversity

Beta diversity measures how different samples are from one another.
`df2beta_pcoa()` builds a sample-by-sample distance matrix and ordinates it with
PCoA. It supports three backends via `method`:

* `"bray"` — Bray-Curtis on proportions (the classic default);
* `"aitchison"` — compositional / log-ratio distance
  (`robCompositions::aDist()`, with zero replacement);
* `"adiv"` — Jaccard-family dissimilarity (`adiv::Jac()`).

We colour points by `stage` (column 7) and draw grouping ellipses.

```{r beta-bray, fig.height = 5}
df_g <- df[df$clade == "G", ]
df2beta_pcoa(df_g, method = "bray", add_legend = 7, add_ellipse = 7)
```

```{r beta-aitchison, eval = requireNamespace("robCompositions", quietly = TRUE), fig.height = 5}
df2beta_pcoa(df_g, method = "aitchison", add_legend = 7, add_ellipse = 7)
```

```{r beta-adiv, eval = requireNamespace("adiv", quietly = TRUE), fig.height = 5}
df2beta_pcoa(df_g, method = "adiv", add_legend = 7, add_ellipse = 7)
```

## 5.1 Statistics: PERMANOVA

`df2beta(..., print_df = TRUE)` returns the raw distance matrix, which we can
feed to `vegan::adonis2()` for a PERMANOVA test of whether stage explains a
significant fraction of the between-sample variance.

```{r beta-dist}
d_bray <- stats::as.dist(df2beta(df_g, method = "bray", print_df = TRUE))
grp <- meta$stage[match(labels(d_bray), meta$sample)]
```

```{r beta-permanova, eval = requireNamespace("vegan", quietly = TRUE)}
set.seed(1)
vegan::adonis2(d_bray ~ grp)
```

## 5.2 A polished PCoA scatter

For full control over styling we can ordinate the Bray-Curtis distance
ourselves with `ape::pcoa()` and draw it with `ggpubr::ggscatter()`, adding
confidence ellipses and group mean points in viridis.

```{r beta-ggscatter, eval = requireNamespace("ggpubr", quietly = TRUE), fig.height = 5}
pcoa <- ape::pcoa(as.matrix(df2beta(df_g, method = "bray", print_df = TRUE)))
scores <- as.data.frame(pcoa$vectors[, 1:2])
names(scores) <- c("PCoA1", "PCoA2")
scores$sample <- rownames(scores)
scores <- dplyr::left_join(scores, meta, by = "sample")

ggpubr::ggscatter(
  scores, x = "PCoA1", y = "PCoA2",
  color = "stage", palette = viridis::viridis(2),
  ellipse = TRUE, ellipse.type = "norm", mean.point = TRUE,
  size = 3, alpha = 0.85
) +
  labs(title = "Bray-Curtis PCoA")
```

# 6. Ordination, clustering and correlation

The matrix-oriented plots operate on a taxa-by-sample matrix such as `mat_g`.

`df2pca_sample()` runs a PCA of the samples. Encoding the legend into the column
names lets us colour the individuals by stage.

```{r pca-sample, fig.height = 5}
mat_g_enc <- df_untidy(df, clade = "G", top = 12, keep_sample_name = FALSE)
df2pca_sample(mat_g_enc, detect = "pupa", detect2 = "larvae") +
  scale_color_viridis_d(end = 0.85)
```

`df2heatmap()` draws a base-graphics heatmap (viridis colour scale)...

```{r heatmap, fig.height = 6}
df2heatmap(mat_g, scale = "row", Colv = NA)
```

...and `df2corrplot()` a clustered taxa-by-taxa correlation plot.

```{r corrplot, fig.height = 6}
df2corrplot(mat_g, k_means = 3)
```

t-SNE (`df2tsne()`) is available for larger taxa sets. It is iterative and can
be slow, so we show the call but leave it un-evaluated in this quick build.

```{r tsne, eval = FALSE}
mat_tsne <- df_untidy(df, clade = "G", top = 40, scale = "scale")
set.seed(1)
df2tsne(mat_tsne, k_means = 4, text_top = 5, perplexity = 5, max_iter = 300)
```

# 7. Differential abundance

`df2volcano()` runs a per-taxon *t*-test between two sample groups and plots the
log abundance change against the p-value. Here we contrast pupa with larvae at
the genus level.

```{r volcano, fig.height = 5}
df2volcano(df_g, legend_detect = c("pupa", "larvae"))
```

# 8. Co-occurrence networks

`df2graph()` turns a taxa-by-sample matrix into an annotated
`tidygraph::tbl_graph` (node centralities + community membership, signed
correlation edges) and `df2ggraph()` draws it with `ggraph` — communities in
viridis, edges coloured by the sign of the correlation.

```{r network, fig.height = 6}
mat_net <- df_untidy(df, clade = "G", top = 15, scale = "scale")
set.seed(1)
df2ggraph(mat_net, threshold = 0.3)
```

## 8.1 NetCoMi networks (optional)

For association networks with dedicated microbiome methods (SparCC, SPIEC-EASI,
SPRING, ...), `df2netcomi()` wraps [`NetCoMi`](https://github.com/stefpeschel/NetCoMi),
and `df2netcomi_graph()` bridges the result into the *same* `df2ggraph()` engine
for a consistent look. `NetCoMi` is a heavy optional dependency, so the chunk
below is shown for reference but not evaluated; install `NetCoMi` and set
`eval = TRUE` to run it.

```{r netcomi, eval = FALSE}
res <- df2netcomi(mat_net, measure = "pearson", thresh = 0.3)
res$props                                 # network summary statistics
df2netcomi_graph(res, plot = TRUE)        # drawn with df2ggraph()
```

# Session info

```{r session}
sessionInfo()
```
