---
title: "Working from a phyloseq object"
author: "aRchiteutis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Working from a phyloseq object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 100,
  dev = "png"
)

Sys.setenv(RGL_USE_NULL = "TRUE")
options(rgl.useNULL = TRUE)
set.seed(1)

library(aRchiteutis)
library(ggplot2)

# The entire vignette needs phyloseq; when it is missing every code chunk is
# skipped so the package still builds (R CMD check robustness).
have_ps <- requireNamespace("phyloseq", quietly = TRUE)
knitr::opts_chunk$set(eval = have_ps)

have_ggpubr <- requireNamespace("ggpubr", quietly = TRUE)
if (have_ggpubr) {
  ggplot2::theme_set(ggpubr::theme_pubr())
} else {
  ggplot2::theme_set(ggplot2::theme_light())
}
```

`aRchiteutis` is built around a single long (tidy) count table, but every
`df2*` / `df_*` function *also* accepts a [`phyloseq`](https://joey711.github.io/phyloseq/)
object directly, thanks to two bridges:

* `ps2df()` — turn a `phyloseq` object into the long table, reproducing the
  multi-rank clade structure that `get_counts()` produces for Kraken2 reports;
* `df2ps()` — the reverse conversion.

This vignette proves that interoperability end-to-end: we build a `phyloseq`
object, round-trip it, and then run a representative slice of the same pipeline
(composition, alpha diversity + statistics, beta diversity with the
compositional backend, and a co-occurrence network) either on the `phyloseq`
object itself or on its `ps2df()` conversion. Styling matches the getting-started
vignette: a light `ggpubr` theme, viridis colours and `ggpubr` statistical
layers.

```{r not-installed, eval = !have_ps, echo = FALSE, results = "asis"}
cat("> **Note:** the `phyloseq` package is not installed, so the code chunks",
    "in this vignette are not evaluated. Install it with",
    "`BiocManager::install(\"phyloseq\")` to run the workflow.\n")
```

# 1. Build a phyloseq object and round-trip it

We start from the honey-bee example data shipped with the package and turn it
into a `phyloseq` object with `df2ps()`. Using the shipped data (rather than a
`phyloseq` built-in such as `GlobalPatterns`) keeps the `stage` / `hive` sample
metadata, which we need for the statistical comparisons below.

```{r build-ps}
path <- system.file("extdata", package = "aRchiteutis")
legend <- system.file("extdata", "legend.csv", package = "aRchiteutis")

df <- get_counts(
  path = path,
  pattern = "_decont_bacteria_k2",
  legend = legend,
  trim_char = "_"
)

ps <- df2ps(df)          # species-level operational taxa + full lineage
ps
```

`ps2df()` converts it straight back into the long table the rest of the package
consumes; the genus read counts survive the round-trip unchanged, a convenient
correctness check.

```{r roundtrip}
psdf <- ps2df(ps)
head(psdf)

# Genus counts before vs after the round-trip match.
before <- sort(tapply(df$N[df$clade == "G"], df$taxa[df$clade == "G"], sum))
after  <- sort(tapply(psdf$N[psdf$clade == "G"], psdf$taxa[psdf$clade == "G"], sum))
all.equal(before, after)
```

The `sample_data` becomes the legend columns (`stage` in column 7, `hive` in
column 8), exactly as `get_counts(legend = ...)` would produce.

```{r ps-meta}
meta <- unique(psdf[, c("sample", "stage", "hive")])
meta
phyloseq::sample_data(ps)
```

# 2. Composition

`ps2df()` output flows into the composition plots just like native data. We trim
to the top genera first (rolling the rest into an "other" group) and draw the
per-sample stacked bars in viridis.

```{r ps-composition, fig.height = 5}
psdf_top <- df_taxa_trim(psdf[psdf$clade != "S", ], top_taxa = 8)
df2composition(psdf_top)
```

```{r ps-donut, fig.height = 5}
df2donut(psdf_top)
```

# 3. Alpha diversity + statistics

We compute per-sample Shannon and Simpson indices from the species-level rows of
the converted table and compare stages with a Wilcoxon test annotated by
`ggpubr::stat_compare_means()` — viridis fills, light theme.

```{r ps-alpha-table}
psdf_sp <- psdf[psdf$clade == "S", ]
alpha_tab <- psdf_sp %>%
  dplyr::group_by(sample) %>%
  dplyr::summarise(
    Shannon = abdiv::shannon(N),
    Simpson = abdiv::simpson(N),
    .groups = "drop"
  )
alpha_tab <- dplyr::left_join(alpha_tab, meta, by = "sample")
alpha_long <- tidyr::pivot_longer(alpha_tab, c("Shannon", "Simpson"),
                                  names_to = "index", values_to = "value")
head(alpha_tab)
```

```{r ps-alpha-stats, eval = have_ps && requireNamespace("ggpubr", quietly = TRUE), fig.height = 4.5}
ggpubr::ggboxplot(
  alpha_long, x = "stage", y = "value", fill = "stage",
  palette = viridis::viridis(2), alpha = 0.7,
  add = "jitter", add.params = list(color = "grey30", size = 1.5),
  facet.by = "index", scales = "free_y"
) +
  ggpubr::stat_compare_means(
    ggplot2::aes(label = ggplot2::after_stat(p.format)),
    method = "wilcox.test"
  ) +
  labs(x = NULL, y = "Diversity index")
```

# 4. Beta diversity directly on the phyloseq object

Because the beta-diversity functions call `as_samovar_df()` / `as_samovar_matrix()`
internally, we can hand them the `phyloseq` object **directly**. `df2beta()`
defaults to the genus clade, so `df2beta(ps, print_df = TRUE)` returns a
genus-level Bray-Curtis distance matrix with no manual conversion.

```{r ps-beta-dist}
d <- stats::as.dist(df2beta(ps, print_df = TRUE))
grp <- meta$stage[match(labels(d), meta$sample)]
as.matrix(d)[1:4, 1:4]
```

```{r ps-permanova, eval = have_ps && requireNamespace("vegan", quietly = TRUE)}
set.seed(1)
vegan::adonis2(d ~ grp)
```

A PCoA ordination, coloured by stage with grouping ellipses. Here we use the
compositional (Aitchison) backend on the converted genus table when
`robCompositions` is available, falling back to Bray-Curtis otherwise.

```{r ps-pcoa, fig.height = 5}
psdf_g <- psdf[psdf$clade == "G", ]
beta_method <- if (requireNamespace("robCompositions", quietly = TRUE)) {
  "aitchison"
} else {
  "bray"
}
df2beta_pcoa(psdf_g, method = beta_method, add_legend = 7, add_ellipse = 7)
```

A polished version with `ggpubr::ggscatter()` (confidence ellipses, group means,
viridis):

```{r ps-ggscatter, eval = have_ps && requireNamespace("ggpubr", quietly = TRUE), fig.height = 5}
pcoa <- ape::pcoa(as.matrix(df2beta(ps, print_df = TRUE)))
scores <- as.data.frame(pcoa$vectors[, 1:2])
names(scores) <- c("PCoA1", "PCoA2")
scores$sample <- rownames(scores)
scores <- dplyr::left_join(scores, meta, by = "sample")

ggpubr::ggscatter(
  scores, x = "PCoA1", y = "PCoA2",
  color = "stage", palette = viridis::viridis(2),
  ellipse = TRUE, ellipse.type = "norm", mean.point = TRUE,
  size = 3, alpha = 0.85
) +
  labs(title = "Genus-level PCoA from a phyloseq object")
```

# 5. Co-occurrence network directly on the phyloseq object

`df2graph()` / `df2ggraph()` also accept the `phyloseq` object directly (genus
clade, top taxa by abundance), building the annotated `tidygraph` network and
drawing it with the shared `ggraph` engine — communities in viridis.

```{r ps-network, fig.height = 6}
set.seed(1)
df2ggraph(ps, top = 15, threshold = 0.3)
```

For NetCoMi-based association networks the `phyloseq` object works the same way
via `df2netcomi()` / `df2netcomi_graph()`; that path needs the heavy optional
`NetCoMi` dependency, so it is shown but not evaluated here.

```{r ps-netcomi, eval = FALSE}
res <- df2netcomi(ps, top = 15, measure = "pearson", thresh = 0.3)
df2netcomi_graph(res, plot = TRUE)
```

# Session info

```{r session}
sessionInfo()
```
